---
title: "belief_tutorial"
author: "Chi Zhang"
date: "2025-12-14"
output: html_document
---

# Setup and Package/dataset Loading

```{r setup, message=FALSE, warning=FALSE}
# Load packages
source("scripts/00_packages.R")
```

# Theoretical Hypotheses and Model Formulas

```{r}
# 1. Unidimensional IRT (UIRT)
# Hypothesis 1: All items measure a single latent trait (General Equity Belief)
bf_uni <- bf(
  response ~ 1 + (1 | item) + (1 | id)
)

# 2. Correlated-Traits MIRT (CT-MIRT)
# Hypothesis 2: Three distinct but correlated beliefs (Math, Learner, Pedagogy)
# (0 + ... | id) allows correlations between dimensions
bf_ct <- bf(
  response ~ 1 + (1 | item) + 
    (0 + dim_m + dim_l + dim_p | id)
)

# 3. Bi-factor MIRT (BF-MIRT)
# Hypothesis 3: General belief + Orthogonal specific beliefs
# (||) enforces orthogonality (independence) between general and specific factors
bf_bf <- bf(
  response ~ 1 + (1 | item) + 
    (1 | id) +                
    (0 + dim_m || id) +       
    (0 + dim_l || id) +       
    (0 + dim_p || id)
)
```

# Preprocessing Step 1: Inspection of Original Scale

```{r}
# Load the original 5-point scale data
data_1 <- readRDS("data/processed/data_1.rds")

# Check response distribution
print(table(data_1$response))

# Run the plotting script to generate the descriptive plot
# This script creates the ggplot object 'fig_3'
source("scripts/03_fig3.R")

# Display the plot
fig_3
```


## Baseline null_rsm model
```{r}
# Since we use the same priors for uni_irt models, define it once.
priors_uni <- prior("normal(0, 3)", class = "sd", group = "id") +
  prior("normal(0, 3)", class = "sd", group = "item")

```


```{r}
# Fit RSM on the original 5-point data
null_rsm <- brm(
  formula = bf_uni,
  data = data_1,
  family = brmsfamily("acat", "logit"),
  prior = priors_uni,
  chains = 4, 
  cores = 4, 
  iter = 4000, 
  warmup = 2000,
  seed = 1234,
  file = "models/null_rsm"
)

null_rsm <- add_criterion(
  null_rsm, 
  criterion = c("loo", "waic")
)

saveRDS(null_rsm, "models/null_rsm.rds")
```
### Summary 
```{r}
summary (null_rsm)
```


# Fitting the Collapsed Model (Alternative)


```{r}
# Load the collapsed 4-point scale data
data_2 <- readRDS("data/processed/data_2.rds")

# Fit RSM on the collapsed 4-point data
merge_rsm <- brm(
  formula = bf_uni,
  data = data_2,
  family = brmsfamily("acat", "logit"),
  prior = priors_uni,
  chains = 4, 
  cores = 4,
  iter = 4000, 
  warmup = 2000,
  file = "models/merge_rsm"
)

merge_rsm <- add_criterion(
  merge_rsm, 
  criterion = c("loo", "waic")
)

saveRDS(merge_rsm, "models/merge_rsm.rds")
```
## Output and comparasion
```{r}
loo (null_rsm,merge_rsm) # merge_rsm is way better!
```
```{r}
source("scripts/04_fig4.R")

fig_4
```


```{r}
summary (merge_rsm) # but still, the gaping issue remains.
```

# Phase 3: Dimensionality Verification

We now test the competing dimensional structures using the collapsed dataset (`data_2`). 
Note that `merge_rsm` already serves as our UIRT baseline for this comparison.

## Fitting the Correlated-Traits Model (CT-MIRT)

For the Correlated-Traits model, we need to specify a prior for the correlation matrix between the three latent traits. We use an `LKJ(2)` prior, which weakly favors lower correlations, aiding convergence.

Given the increased model complexity, we also increase `max_treedepth` to 15 to prevent NUTS saturation.

```{r}
# Define priors for CT model: Base priors + Correlation prior
priors_ct <- priors_uni + prior(lkj(2), class = "cor")

# Fit CT-MIRT
ct_rsm <- brm(
  formula = bf_ct,
  data = data_2,
  family = brmsfamily("acat", "logit"),
  prior = priors_ct,
  chains = 4, 
  cores = 4, 
  iter = 4000, 
  warmup = 2000, 
  seed = 1234,
  # Tutorial Adjustment: Increase treedepth for complex MIRT
  control = list(adapt_delta = 0.99, max_treedepth = 15), 
  file = "models/ct_rsm"
)

ct_rsm <- add_criterion(
  ct_rsm, 
  criterion = c("loo", "waic")
)

saveRDS(ct_rsm, "models/ct_rsm.rds")
```

```{r}
summary (ct_rsm)
```

## Fitting the Bi-factor Model (BF-MIRT)
The Bi-factor model assumes orthogonal specific factors. While we don't need a correlation prior, the orthogonality constraints make the posterior geometry difficult to explore. We maintain the robust control settings.

```{r}
# Fit BF-MIRT
# Note: bf_bf formula uses '||' which assumes orthogonality (no correlations)
bf_rsm <- brm(
  formula = bf_bf,
  data = data_2,
  family = brmsfamily("acat", "logit"),
  prior = priors_uni, # Reuse standard SD priors
  chains = 4, 
  cores = 4, 
  iter = 4000, 
  warmup = 2000, 
  seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_rsm"
)

bf_rsm <- add_criterion(
  bf_rsm, 
  criterion = c("loo", "waic")
)

saveRDS(bf_rsm, "models/bf_rsm.rds")
```


## Model Comparison: Dimensionality
We compare the three structures (Uni, CT, BF) to determine the best-fitting dimensionality for the equity beliefs.

```{r}
loo_compare(merge_rsm, ct_rsm, bf_rsm, criterion = "loo")
```
# Discrimination check

```{r}
# CT-2pl Formula
bf_ct_2pl <- bf(
  response ~ 1 + (1 | item) + (0 + dim_m + dim_l + dim_p | id),
  disc ~ 1 + (1 | item)
)

priors_ct_2pl <- priors_ct + 
  prior("normal(0, 1)", class = "sd", group = "item", dpar = "disc")

# BF-2pl Formula
bf_bf_2pl <- bf(
  response ~ 1 + (1 | item) + 
    (1 | id) +                
    (0 + dim_m || id) +       
    (0 + dim_l || id) +       
    (0 + dim_p || id),        
  disc ~ 1 + (1 | item)       
)

priors_bf_2pl <- priors_uni + 
  prior("normal(0, 1)", class = "sd", group = "item", dpar = "disc")
```

## Fit 2pl (ct_grsm, bf_grsm)
```{r}

ct_grsm <- brm(
  formula = bf_ct_2pl,
  data = data_2,
  family = brmsfamily("acat", "logit"),
  prior = priors_ct_2pl,
  chains = 4, 
  cores = 4, 
  iter = 4000, 
  warmup = 2000, 
  seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/ct_grsm"
)

ct_grsm <- add_criterion(ct_grsm, criterion = c("loo", "waic"), overwrite = TRUE)
saveRDS(ct_grsm, "models/ct_grsm.rds")

bf_grsm <- brm(
  formula = bf_bf_2pl,
  data = data_2,
  family = brmsfamily("acat", "logit"),
  prior = priors_bf_2pl,
  chains = 4, 
  cores = 4, 
  iter = 4000, 
  warmup = 2000, 
  seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_grsm"
)

bf_grsm <- add_criterion(bf_grsm, criterion = c("loo", "waic"), overwrite = TRUE)
saveRDS(bf_grsm, "models/bf_grsm.rds")
```

## Output and compare

```{r}
loo_compare(ct_rsm, bf_rsm, ct_grsm, bf_grsm, criterion = "loo")
```

```{r}
summary (bf_grsm)
```

```{r}

source ("scripts/05_fig5.R")
fig_5

```

# Delete poor items

```{r}
# This provides data3 that is without bm1r~bm4r
data_3 <- readRDS("data/processed/data_3.rds")
```

# Model setting compare

```{r}

# --- 1. GRSM (Family: acat) on Data 3 ---

# CT-GRSM (New)
ct_grsm_new <- brm(
  formula = bf_ct_2pl,
  data = data_3,
  family = brmsfamily("acat", "logit"),
  prior = priors_ct_2pl,
  backend = "cmdstanr",  
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/ct_grsm_new"
)
ct_grsm_new <- add_criterion(ct_grsm_new, "loo", overwrite = TRUE)
saveRDS(ct_grsm_new, "models/ct_grsm_new.rds")

# BF-GRSM (New)
bf_grsm_new <- brm(
  formula = bf_bf_2pl,
  data = data_3,
  family = brmsfamily("acat", "logit"),
  prior = priors_bf_2pl,
  backend = "cmdstanr",  
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_grsm_new"
)
bf_grsm_new <- add_criterion(bf_grsm_new, "loo")


# --- 2. GRM (Family: cumulative) on Data 3 ---

# CT-GRM (New)
ct_grm_new <- brm(
  formula = bf_ct_2pl,
  data = data_3,
  family = brmsfamily("cumulative", "logit"), # Changed to cumulative for grm setting
  prior = priors_ct_2pl,
  backend = "cmdstanr", 
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/ct_grm_new"
)
ct_grm_new <- add_criterion(ct_grm_new, "loo", overwrite = TRUE)
saveRDS(ct_grm_new, "models/ct_grm_new.rds")

# BF-GRM (New)
bf_grm_new <- brm(
  formula = bf_bf_2pl,
  data = data_3,
  family = brmsfamily("cumulative", "logit"), 
  prior = priors_bf_2pl,
  backend = "cmdstanr", 
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_grm_new"
)
bf_grm_new <- add_criterion(bf_grm_new, "loo", overwrite = TRUE)
saveRDS(bf_grm_new, "models/bf_grm_new.rds")

# --- 3. GPCM on Data 3 ---
# Formula: Needs cs(1) to allow thresholds to vary by item (Partial Credit)

# Define PCM Formulas
bf_ct_gpcm <- bf(
  response ~ 1 + (cs(1) | item) + (0 + dim_m + dim_l + dim_p | id),
  disc ~ 1 + (1 | item)
)

bf_bf_gpcm <- bf(
  response ~ 1 + (cs(1) | item) + 
    (1 | id) + (0 + dim_m || id) + (0 + dim_l || id) + (0 + dim_p || id),
  disc ~ 1 + (1 | item)
)

# Fit CT-PCM (New)
ct_gpcm_new <- brm(
  formula = bf_ct_gpcm,
  data = data_3,
  family = brmsfamily("acat", "logit"),
  prior = priors_ct_2pl, 
  backend = "cmdstanr",  
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/ct_gpcm_new"
)
ct_gpcm_new <- add_criterion(ct_gpcm_new, "loo", overwrite = TRUE)
saveRDS(ct_gpcm_new, "models/ct_gpcm_new.rds")

# Fit BF-GPCM 
bf_gpcm_new <- brm(
  formula = bf_bf_gpcm,
  data = data_3,
  family = brmsfamily("acat", "logit"),
  prior = priors_bf_2pl, 
  backend = "cmdstanr",  
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_gpcm_new"
)
bf_gpcm_new <- add_criterion(bf_gpcm_new, "loo", overwrite = TRUE)
saveRDS(bf_gpcm_new, "models/bf_gpcm_new.rds")
```


```{r}
# Final 6-Model Comparison on Data 3
loo_compare(
  ct_grsm_new, bf_grsm_new, 
  ct_grm_new,  bf_grm_new,
  ct_gpcm_new,  bf_gpcm_new,
  criterion = "loo"
)

model_weights (ct_grm_new,  bf_grm_new)
```

# Summary for final choice
```{r}
source("scripts/07_fig6.R")

fig_6
```


# DIF models
```{r}
bf_dif <- bf(
  response ~ 1 + gender + (0 + gender | item) + 
    (1 | id) + (0 + dim_m || id) + (0 + dim_l || id) + (0 + dim_p || id),
  disc ~ 1 + (1 | item)
)

bf_grm_dif <- brm(
  formula = bf_dif,
  data = data_3,
  family = brmsfamily("cumulative", "logit"),
  prior = priors_bf_2pl,  
  backend = "cmdstanr",
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_grm_dif"
)

```



```{r}

bf_dif2 <- bf(
  response ~ 1 + gender + District + YearG + 
    (1 + gender + District | item) + 
    (1 | id) + (0 + dim_m || id) + (0 + dim_l || id) + (0 + dim_p || id),
  disc ~ 1 + (1 | item)
)

# Fit Model
bf_grm_dif2 <- brm(
  formula = bf_dif2,
  data = data_3,
  family = brmsfamily("cumulative", "logit"),
  prior = priors_bf_2pl, 
  backend = "cmdstanr",
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_grm_dif2"
)
```


