---
title: "belief_tutorial"
author: "Chi Zhang & Maria Pampaka"
date: "2026-03-21"
output: html_document
---

# Setup and Package/dataset Loading

```{r setup, message=FALSE, warning=FALSE}
# Load packages
source("scripts/00_packages.R")
```

# Model Formulas: Three Competing Dimensional Structures
In brms, IRT models are specified as generalized linear mixed models 
(GLMMs). Item parameters are modelled as random intercepts grouped by 
item, and person parameters as random effects grouped by person id. The 
three structural hypotheses translate directly into different random 
effect specifications:

```{r}
# 1. Unidimensional IRT (UIRT)
# All items load on a single latent trait.
# (1 | item) = item-specific difficulty (random intercept)
# (1 | id)   = person ability (random intercept)
bf_uni <- bf(
  response ~ 1 + (1 | item) + (1 | id)
)

# 2. Correlated-Traits MIRT (CT-MIRT)
# Items load on one of three correlated dimensions.
# dim_m, dim_l, dim_p are binary indicators (0/1) assigning each item 
# to its dimension.
# (0 + dim_m + dim_l + dim_p | id) estimates three person-level traits 
# with an unrestricted covariance matrix, allowing correlations between 
# dimensions.
bf_ct <- bf(
  response ~ 1 + (1 | item) + 
    (0 + dim_m + dim_l + dim_p | id)
)

# 3. Bi-factor MIRT (BF-MIRT)
# A general factor (1 | id) plus dimension-specific factors.
# The double-bar notation (||) in brms suppresses correlations between 
# random effects.
bf_bf <- bf(
  response ~ 1 + (1 | item) + 
    (1 || id) +
    (0 + dim_m || id) +
    (0 + dim_l || id) +
    (0 + dim_p || id)
)
```

# Prior Specification

We use weakly informative priors throughout, following recommendations 
in Bürkner (2021). These provide sufficient regularisation to aid 
convergence without imposing strong constraints on parameter estimates.

```{r priors}
# Standard priors for person and item random effects
priors_uni <- prior("normal(0, 3)", class = "sd", group = "id") +
  prior("normal(0, 3)", class = "sd", group = "item")

# For CT models: add LKJ(2) prior on the correlation matrix between 
# latent dimensions. LKJ with eta > 1 weakly favours smaller 
# correlations, aiding convergence.
priors_ct <- priors_uni + prior(lkj(2), class = "cor")

# For 2PL models: add prior for log-discrimination.
priors_ct_2pl <- priors_ct +
  prior("normal(0, 1)", class = "sd", group = "item", dpar = "disc")

priors_bf_2pl <- priors_uni +
  prior("normal(0, 1)", class = "sd", group = "item", dpar = "disc")
```


# Step 1: Preprocessing sparse response categories

## Inspecting the original 5-point scale

```{r}
# Load the original 5-point scale data
data_1 <- readRDS("data/processed/data_1.rds")

# Check response distribution
print(table(data_1$response))

# Run the plotting script to generate the descriptive plot
# This script creates the ggplot object 'fig_3'
source("scripts/03_fig3.R")

fig_3
```


## Baseline null_rsm model

```{r}
# Fit RSM on the original 5-point data
null_rsm <- brm(
  formula = bf_uni,
  data = data_1,
  family = brmsfamily("acat", "logit"),
  prior = priors_uni,
  chains = 4, 
  cores = 4, 
  iter = 4000, 
  warmup = 2000,
  seed = 1234,
  file = "models/null_rsm"
)

null_rsm <- add_criterion(
  null_rsm, 
  criterion = c("loo", "waic")
)

saveRDS(null_rsm, "models/null_rsm.rds")
```

### Summary 
```{r}
summary (null_rsm)
```

## Collapsing sparse categories

Twelve items had fewer than three responses in the lowest category 
("strongly disagree"), providing insufficient information for reliable 
threshold estimation. We collapse categories 1 and 2 into a single 
category, reducing the scale from 5 to 4 ordered categories.


```{r}
# Load the collapsed 4-point scale data
data_2 <- readRDS("data/processed/data_2.rds")

# Fit RSM on the collapsed 4-point data
merge_rsm <- brm(
  formula = bf_uni,
  data = data_2,
  family = brmsfamily("acat", "logit"),
  prior = priors_uni,
  chains = 4, 
  cores = 4,
  iter = 4000, 
  warmup = 2000,
  file = "models/merge_rsm"
)

merge_rsm <- add_criterion(
  merge_rsm, 
  criterion = c("loo", "waic")
)

saveRDS(merge_rsm, "models/merge_rsm.rds")
```

```{r}
summary(merge_rsm)
```
### A note on comparing models across different response scales

```{r}
# You might attempt to compare the two models using LOO-CV:
loo(null_rsm, merge_rsm)

# However, this comparison should be interpreted with caution. 
# The two models predict different response variables (5 categories 
# vs 4 categories). Reducing the number of categories mechanically 
# increases per-observation predictive density — predicting among 
# 4 outcomes is inherently "easier" than predicting among 5. 
#
# The decision to collapse categories is instead justified by:
# (1) Descriptive evidence: 12 items had < 3 responses in category 1.
# (2) Diagnostic evidence: the initial RSM produced anomalous threshold
#     ordering (see summary above), where the model could not 
#     meaningfully distinguish between the first two category transitions.
```
### Summary

```{r}
source("scripts/04_fig4.R")

fig_4
```

```{r}
summary (merge_rsm) # but still, the gaping issue remains.
```

# Step 2: Dimensionality — Unidimensional vs Multidimensional

We compare the three competing dimensional structures using the 
collapsed dataset (`data_2`). All models here use the same 1PL (RSM) 
specification with the same 4-category response variable, making LOO-CV 
comparisons valid.

## Fitting the Correlated-Traits model



```{r}
# Define priors for CT model: Base priors + Correlation prior
priors_ct <- priors_uni + prior(lkj(2), class = "cor")

# Fit CT-MIRT
ct_rsm <- brm(
  formula = bf_ct,
  data = data_2,
  family = brmsfamily("acat", "logit"),
  prior = priors_ct,
  chains = 4, 
  cores = 4, 
  iter = 4000, 
  warmup = 2000, 
  seed = 1234,
  # Increase treedepth for complex MIRT
  control = list(adapt_delta = 0.99, max_treedepth = 15), 
  file = "models/ct_rsm"
)

ct_rsm <- add_criterion(
  ct_rsm, 
  criterion = c("loo", "waic")
)

saveRDS(ct_rsm, "models/ct_rsm.rds")
```



## Fitting the Bi-factor Model (BF-MIRT)

```{r}
bf_rsm <- brm(
  formula = bf_bf,
  data = data_2,
  family = brmsfamily("acat", "logit"),
  prior = priors_uni, 
  chains = 4, 
  cores = 4, 
  iter = 4000, 
  warmup = 2000, 
  seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_rsm"
)

bf_rsm <- add_criterion(
  bf_rsm, 
  criterion = c("loo", "waic")
)

saveRDS(bf_rsm, "models/bf_rsm.rds")
```



```{r}
summary (ct_rsm)
summary (bf_rsm)
```
## Dimensionality comparison

```{r}
loo_compare(merge_rsm, ct_rsm, bf_rsm, criterion = "loo")
# Both CT and BF substantially outperform the unidimensional model,
# confirming that the data have multidimensional structure.
# However, CT and BF are not clearly distinguishable at this stage.
```
# Step 3: Freeing Discrimination Parameters
The RSM constrains all item discriminations to be equal. When items 
differ substantially in how well they differentiate between respondents, 
this constraint forces the model to find compromise threshold values, 
which can produce anomalous threshold ordering. Freeing discrimination 
parameters (moving from 1PL to 2PL) serves a dual diagnostic purpose: 
it tests the equal-discrimination assumption and reveals items with 
poor measurement properties.

## 2PL formula specification
```{r}
# CT-2pl Formula
bf_ct_2pl <- bf(
  response ~ 1 + (1 | item) + (0 + dim_m + dim_l + dim_p | id),
  disc ~ 1 + (1 | item)
)

priors_ct_2pl <- priors_ct + 
  prior("normal(0, 1)", class = "sd", group = "item", dpar = "disc")

# BF-2pl Formula
bf_bf_2pl <- bf(
  response ~ 1 + (1 | item) + 
    (1 | id) +                
    (0 + dim_m || id) +       
    (0 + dim_l || id) +       
    (0 + dim_p || id),        
  disc ~ 1 + (1 | item)       
)

priors_bf_2pl <- priors_uni + 
  prior("normal(0, 1)", class = "sd", group = "item", dpar = "disc")
```

## Fit 2pl (ct_grsm, bf_grsm)
```{r}

ct_grsm <- brm(
  formula = bf_ct_2pl,
  data = data_2,
  family = brmsfamily("acat", "logit"),
  prior = priors_ct_2pl,
  chains = 4, 
  cores = 4, 
  iter = 4000, 
  warmup = 2000, 
  seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/ct_grsm"
)

ct_grsm <- add_criterion(ct_grsm, criterion = c("loo", "waic"), overwrite = TRUE)
saveRDS(ct_grsm, "models/ct_grsm.rds")

bf_grsm <- brm(
  formula = bf_bf_2pl,
  data = data_2,
  family = brmsfamily("acat", "logit"),
  prior = priors_bf_2pl,
  chains = 4, 
  cores = 4, 
  iter = 4000, 
  warmup = 2000, 
  seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_grsm"
)

bf_grsm <- add_criterion(bf_grsm, criterion = c("loo", "waic"), overwrite = TRUE)
saveRDS(bf_grsm, "models/bf_grsm.rds")
```

## Comparing 1PL and 2PL specifications

```{r}
loo_compare(ct_rsm, bf_rsm, ct_grsm, bf_grsm, criterion = "loo")
```

```{r}
summary (bf_grsm)
```

## Wright map: inspecting discrimination and difficulty

```{r}

source ("scripts/05_fig5.R")
fig_5

# The Wright map reveals substantial variation in discrimination 
# across items, confirming that the equal-discrimination assumption 
# was untenable. Threshold ordering has also improved (see paper 
# Table for threshold estimates).
#
# Critically, four items (Bm_1_r to Bm_4_r) stand out with:
# - Discrimination estimates near zero
# - Extremely wide credible intervals on difficulty
# These are candidates for removal.
```

# Step 4: Item removal and different link function comparisons

```{r}
# This provides data3 that is without bm1r~bm4r
data_3 <- readRDS("data/processed/data_3.rds")
```

With dimensionality confirmed and problematic items removed, we now 
compare IRT models defined by different ordinal link functions. 

## Fitting the six models

Three link function families are compared, each crossed with two 
dimensional structures (CT and BF), yielding six models:



```{r}
# --- GRSM: Generalised Rating Scale Model ---
# Link: adjacent-category logit
# Thresholds: shared across items (only item intercepts vary)
# In brms: family = brmsfamily("acat", "logit")
# Formula:  response ~ 1 + (1 | item) + ...
#
# --- GRM: Graded Response Model ---
# Link: cumulative logit
# Thresholds: item-specific (estimated via cumulative family)
# In brms: family = brmsfamily("cumulative", "logit")
# Formula:  same as GRSM; the family handles the parameterisation
#
# --- GPCM: Generalised Partial Credit Model ---
# Link: adjacent-category logit
# Thresholds: item-specific (via category-specific effects)
# In brms: family = brmsfamily("acat", "logit")
# Formula:  response ~ 1 + (cs(1) | item) + ...
#           cs(1) allows thresholds to vary by item


# --- 1. GRSM  ---

ct_grsm_20 <- brm(
  formula = bf_ct_2pl,
  data = data_3,
  family = brmsfamily("acat", "logit"),
  prior = priors_ct_2pl,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/ct_grsm_20"
)
ct_grsm_20 <- add_criterion(ct_grsm_20, "loo", overwrite = TRUE)
saveRDS(ct_grsm_20, "models/ct_grsm_20.rds")

bf_grsm_20 <- brm(
  formula = bf_bf_2pl,
  data = data_3,
  family = brmsfamily("acat", "logit"),
  prior = priors_bf_2pl,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_grsm_20"
)
bf_grsm_20 <- add_criterion(bf_grsm_20, "loo", overwrite = TRUE)
saveRDS(bf_grsm_20, "models/bf_grsm_20.rds")


# --- 2. GRM  ---

ct_grm_20 <- brm(
  formula = bf_ct_2pl,
  data = data_3,
  family = brmsfamily("cumulative", "logit"),
  prior = priors_ct_2pl,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/ct_grm_20"
)
ct_grm_20 <- add_criterion(ct_grm_20, "loo", overwrite = TRUE)
saveRDS(ct_grm_20, "models/ct_grm_20.rds")

bf_grm_20 <- brm(
  formula = bf_bf_2pl,
  data = data_3,
  family = brmsfamily("cumulative", "logit"),
  prior = priors_bf_2pl,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_grm_20"
)
bf_grm_20 <- add_criterion(bf_grm_20, "loo", overwrite = TRUE)
saveRDS(bf_grm_20, "models/bf_grm_20.rds")

# --- 3. GPCM  ---

bf_ct_gpcm <- bf(
  response ~ 1 + (cs(1) | item) + (0 + dim_m + dim_l + dim_p | id),
  disc ~ 1 + (1 | item)
)

bf_bf_gpcm <- bf(
  response ~ 1 + (cs(1) | item) +
    (1 | id) + (0 + dim_m || id) + (0 + dim_l || id) + (0 + dim_p || id),
  disc ~ 1 + (1 | item)
)

ct_gpcm_20 <- brm(
  formula = bf_ct_gpcm,
  data = data_3,
  family = brmsfamily("acat", "logit"),
  prior = priors_ct_2pl,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/ct_gpcm_20"
)
ct_gpcm_20 <- add_criterion(ct_gpcm_20, "loo", overwrite = TRUE)
saveRDS(ct_gpcm_20, "models/ct_gpcm_20.rds")

bf_gpcm_20 <- brm(
  formula = bf_bf_gpcm,
  data = data_3,
  family = brmsfamily("acat", "logit"),
  prior = priors_bf_2pl,
  backend = "cmdstanr",
  save_pars = save_pars(all = TRUE),
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_gpcm_20"
)
bf_gpcm_20 <- add_criterion(bf_gpcm_20, "loo", overwrite = TRUE)
saveRDS(bf_gpcm_20, "models/bf_gpcm_20.rds")
```

## Inspecting individual LOO diagnostics

Individual LOO summaries reveal Pareto-k diagnostics for each model. 
High Pareto-k values (> 0.7) indicate observations where the LOO 
approximation may be unreliable. A model with the best ELPD but many 
high Pareto-k values should be treated with caution.

```{r}
loo(ct_grsm_20)
loo(bf_grsm_20)
loo(ct_grm_20)
loo(bf_grm_20)
loo(ct_gpcm_20)
loo(bf_gpcm_20)
```


```{r}
loo_compare(
  ct_grsm_20, bf_grsm_20, 
  ct_grm_20,  bf_grm_20,
  ct_gpcm_20, bf_gpcm_20
)


# Stacking weights across all six models
model_weights(
  ct_grsm_20, bf_grsm_20,
  ct_grm_20,  bf_grm_20,
  ct_gpcm_20, bf_gpcm_20
)


```


## Clean loo

```{r}


# CT-GRM
loo_ct_grm_20 <- loo(ct_grm_20, moment_match = TRUE, reloo = TRUE)
saveRDS(loo_ct_grm_20, "models/loo/loo_ct_grm_20.rds")

# 4. BF-GRM
loo_bf_grm_20 <- loo(bf_grm_20, moment_match = TRUE)
saveRDS(loo_bf_grm_20, "models/loo/loo_bf_grm_20.rds")


```
## Clean compare

```{r}
loo_compare(loo_bf_grm_20, loo_ct_grm_20)

loo_model_weights(list(bf_grm = loo_bf_grm_20, ct_grm = loo_ct_grm_20))
```



# Step 5: Final model Wright map
```{r}
source("scripts/07_fig6.R")

fig_6
```
```{r}
pp_check(bf_grm_20,type = "bars", ndraws = 100)
pp_check(bf_grm_20, type = "bars_grouped", group = "item", ndraws = 100)
```

# Step 6: Covariates and Differential Item Functioning

DIF analysis tests whether individual items function differently across 
demographic subgroups. In the Bayesian framework, this is implemented 
by adding group variables as covariates with item-level random slopes, 
enabling partial pooling across items.

## DIF model 1: Gender only
```{r}
bf_dif <- bf(
  response ~ 1 + gender + (0 + gender | item) + 
    (1 | id) + (0 + dim_m || id) + (0 + dim_l || id) + (0 + dim_p || id),
  disc ~ 1 + (1 | item)
)

bf_grm_dif <- brm(
  formula = bf_dif,
  data = data_3,
  family = brmsfamily("cumulative", "logit"),
  prior = priors_bf_2pl,  
  backend = "cmdstanr",
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_grm_dif"
)
```

## DIF model 2: Multiple covariates

```{r}

bf_dif2 <- bf(
  response ~ 1 + gender + District +
    (1 + gender + District || item) + 
    (1 | id) + (0 + dim_m || id) + (0 + dim_l || id) + (0 + dim_p || id),
  disc ~ 1 + (1 | item)
)


bf_grm_dif2 <- brm(
  formula = bf_dif2,
  data = data_3,
  family = brmsfamily("cumulative", "logit"),
  prior = priors_bf_2pl, 
  backend = "cmdstanr",
  chains = 4, cores = 4, iter = 4000, warmup = 2000, seed = 1234,
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  file = "models/bf_grm_dif2"
)
```

## DIF results

```{r step6-results}
summary(bf_grm_dif)
summary(bf_grm_dif2)

source("scripts/08_fig7.R")

fig_7
```


